Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-21 paper-filmat.tex
typo

That said, intuitionistic logic still allows for a considerable amount of
freedom, and in many toposes statements are true which are baffling if one has
only received training in mathematics based on classical logic. For instance, on first sight it
looks like the signum function
\[ \operatorname{sgn} : \RR \longrightarrow \RR,\ x \longmapsto \begin{cases}
  -1, & \text{if $x < 0$,} \\
  0, & \text{if $x = 0$,} \\
  1, & \text{if $x > 0$,}
\end{cases} \]
is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
continuous''. However, closer inspection reveals that the signum function
cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
logic is available. The domain of the signum
function is the subset~$\{ x \in \RR \,|\, x < 0 \vee x = 0 \vee x > 0 \}
\subseteq \RR$, and in intuitionistic logic this subset cannot be shown to
coincide with~$\RR$.

@@

A basic observation in computability theory is that there are computational tasks
which are not solvable even for these idealized Turing machines. The paramount
example is the \emph{halting problem}: Given a turing machine~$M$, determine
whether~$M$ terminates (comes to a stop after having carried out a finite
number of computational steps) or not.

@@

\bigskip
\dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the signum and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true and independently due to
Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
straightforward (a textbook reference
is~\cite[Thm.~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
explanation is as follows.

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
such approximations. Hence a function such as the signum function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, do not exist in the \effective topos.

@@

@@

However, the realization of the fiction of infinitesimal numbers in nonstandard analysis crucically rests on a non-principal
ultrafilter, whose existence requires principles which go beyond the means of
Zermelo--Fraenkel set theory~\textsc{zf}.\footnote{A hyperreal number is
represented by an infinite sequence~$(x_0,x_1,x_2,\ldots)$ of ordinary real
numbers. For instance, the sequence~$(1,1,1,\ldots)$ represents the hyperreal
version of the number~$1$, the sequence~$(1,\frac{1}{2},\frac{1}{3},\ldots)$
represents an infinitesimal number and its inverse~$(1,2,3,\ldots)$ represents
a transfinite number.
%
The sequence~$(1,1,1,\ldots)$ is deemed positive, and so is~$(-1,1,1,1,\ldots)$,
which differs from the former only in finitely many places. But
should~$(1,-1,1,-1,\ldots)$ be deemed positive or negative? Whatever the
answer, our decision has consequences for other sequences. For
instance~$(-1,1,-1,1,\ldots)$ should be assigned the opposite sign
and~$(\tan(1),\tan(-1),\tan(1),\tan(-1),\ldots)$ the same.
%
A non-principal ultrafilter is a set-theoretic gadget which fixes all such
decisions once and for all in a coherent manner. Having such an ultrafilter
available, a sequence~$(x_0,x_1,x_2,\ldots)$ is deemed positive if and only if
the set~$\{i \in \NN \,|\, x_i > 0\}$ is part of the ultrafilter.}
Non-principal ultrafilters cannot be described in explicit terms, and
they are also not at all canonical structures: \textsc{zfc} proves that there
are~$2^{2^{\aleph_0}}$ many~\cite{pospisil:ultrafilters}.

@@

However, the dual numbers are severely lacking in other aspects. Firstly, they
do not contain any nilcubic numbers which are not already nilsquare. These are
required in order to extend calculations like~\eqref{eq:derivative} to second
derivatives, as in
\[ (x+\varepsilon)^3 - x^3 = 3x^2\varepsilon + \frac{1}{2!} 6x
\varepsilon^2. \]
Secondly, the dual numbers contain, up to scaling, only a single infinitesimal
number. Further independent infinitesimals are required in order to deal with
functions of several variables, as in
\[ f(x+\varepsilon,y+\varepsilon') - f(x,y) = D_xf(x,y)\varepsilon +
D_yf(x,y)\varepsilon'. \]
Thirdly, and perhaps most importantly, the ring of dual numbers fails to be a
field. The only invertible dual numbers are the numbers of the form~$a +
b\varepsilon$ with~$a$ invertible in the reals; it is not true that any nonzero
dual number is invertible.

@@

\begin{rem}The ring~$\RR^\sim$ of smooth numbers does not coincide with the Cauchy reals,
the Dedekind reals or indeed any well-known construction of the reals
within~$\Zar(\RR)$. This observation explains why~$\RR^\sim$ can satisfy the law of
cancellation even though it is an intuitionistic theorem that the only
nilquadratic number in any flavor of the reals is zero.
\end{rem}

@@

Hence the nilquadratic infinitesimal numbers of synthetic differential geometry
may freely be employed as a convenient fiction when computing derivatives.
Because the theorem on the existence of well-adapted models has a constructive
proof, any proof making use of these infinitesimals may mechanically be
unwound to a (longer and more complex) proof which only refers to the ordinary
reals.

@@

The crucial point is that the translation scheme provided by any topos is sound
with respect to intuitionistic logic. XXX

>>>>

That said, intuitionistic logic still allows for a considerable amount of
freedom, and in many toposes statements are true which are baffling if one has
only received training in mathematics based on classical logic. For instance, on first sight it
looks like the sign function
\[ \operatorname{sgn} : \RR \longrightarrow \RR,\ x \longmapsto \begin{cases}
  -1, & \text{if $x < 0$,} \\
  0, & \text{if $x = 0$,} \\
  1, & \text{if $x > 0$,}
\end{cases} \]
is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
continuous''. However, closer inspection reveals that the sign function
cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
logic is available. The domain of the sign
function is the subset~$\{ x \in \RR \,|\, x < 0 \vee x = 0 \vee x > 0 \}
\subseteq \RR$, and in intuitionistic logic this subset cannot be shown to
coincide with~$\RR$.

@@

A basic observation in computability theory is that there are computational tasks
which are not solvable even for these idealized Turing machines. The paramount
example is the \emph{halting problem}: Given a Turing machine~$M$, determine
whether~$M$ terminates (comes to a stop after having carried out a finite
number of computational steps) or not.

@@

\bigskip
\dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the sign and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true and independently due to
Kreisel--Lacombe--Shoenfield~\cite{kreisel-lacombe-shoenfield:cont} and
Ceĭtin~\cite{ceitin:cont}. A rigorous proof is not entirely
straightforward (a textbook reference
is~\cite[Thm.~9.2.1]{longley-normann:higher-order-computability}), but an intuitive
explanation is as follows.

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, when given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within in a finite amount of time, the machine~$Y$ can only learn finitely many
such approximations. Hence a function such as the sign function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, do not exist in the \effective topos.

@@

% XXX introduction...

@@

However, the realization of the fiction of infinitesimal numbers in nonstandard analysis crucially rests on a non-principal
ultrafilter, whose existence requires principles which go beyond the means of
Zermelo--Fraenkel set theory~\textsc{zf}.\footnote{A hyperreal number is
represented by an infinite sequence~$(x_0,x_1,x_2,\ldots)$ of ordinary real
numbers. For instance, the sequence~$(1,1,1,\ldots)$ represents the hyperreal
version of the number~$1$, the sequence~$(1,\frac{1}{2},\frac{1}{3},\ldots)$
represents an infinitesimal number and its inverse~$(1,2,3,\ldots)$ represents
a transfinite number.
%
The sequence~$(1,1,1,\ldots)$ is deemed positive, and so is~$(-1,1,1,1,\ldots)$,
which differs from the former only in finitely many places. But
should~$(1,-1,1,-1,\ldots)$ be deemed positive or negative? Whatever the
answer, our decision has consequences for other sequences. For
instance~$(-1,1,-1,1,\ldots)$ should be assigned the opposite sign
and~$(\tan(1),\tan(-1),\tan(1),\tan(-1),\ldots)$ the same.
%
A non-principal ultrafilter is a set-theoretic gadget which fixes all such
decisions once and for all in a coherent manner. Having such an ultrafilter
available, a sequence~$(x_0,x_1,x_2,\ldots)$ is deemed positive if and only if
the set~$\{i \in \NN \,|\, x_i > 0\}$ is part of the ultrafilter.}
Non-principal ultrafilters cannot be described in explicit terms, and
they are also not at all canonical structures: \textsc{zfc} proves that there
are~$2^{2^{\aleph_0}}$ many~\cite{pospisil:ultrafilters}.

@@

However, the dual numbers are severely lacking in other aspects. Firstly, they
do not contain any nilcube numbers which are not already nilsquare. These are
required in order to extend calculations like~\eqref{eq:derivative} to second
derivatives, as in
\[ (x+\varepsilon)^3 - x^3 = 3x^2\varepsilon + \frac{1}{2!} 6x
\varepsilon^2. \]
Secondly, the dual numbers contain, up to scaling, only a single infinitesimal
number. Further independent infinitesimals are required in order to deal with
functions of several variables, as in
\[ f(x+\varepsilon,y+\varepsilon') - f(x,y) = D_xf(x,y)\varepsilon +
D_yf(x,y)\varepsilon'. \]
Thirdly, and perhaps most importantly, the ring of dual numbers fails to be a
field. The only invertible dual numbers are the numbers of the form~$a +
b\varepsilon$ with~$a$ invertible in the reals; it is not true that any nonzero
dual number is invertible.

@@

\begin{rem}The ring~$\RR^\sim$ of smooth numbers does not coincide with the Cauchy reals,
the Dedekind reals or indeed any well-known construction of the reals
within~$\Zar(\RR)$. This observation explains why~$\RR^\sim$ can satisfy the law of
cancellation even though it is an intuitionistic theorem that the only
nilsquare number in any flavor of the reals is zero.
\end{rem}

@@

Hence the nilsquare infinitesimal numbers of synthetic differential geometry
may freely be employed as a convenient fiction when computing derivatives.
Because the theorem on the existence of well-adapted models has a constructive
proof, any proof making use of these infinitesimals may mechanically be
unwound to a (longer and more complex) proof which only refers to the ordinary
reals.

@@

The crucial point is that the translation scheme provided by any topos is sound
with respect to intuitionistic logic. Hence, independent of our position
towards toposes as alternative universe, we XXX
